Temporal Continuity and the Judgment of Actual Causation

### 2021-05-05 analysis/statistical_analyses.Rmd
revised entire paper

d1_stats1 <- d1 %>% 
  mutate(ri = factor(case_when(response=="RC_IC" ~  NA_character_,
                        response=="RC" ~     "root",
                        response=="RD_ID" ~  NA_character_,
                        response=="RD" ~     "root",
                        response=="IC" ~     "immediate",
                        response=="ID" ~     "immediate",
                        TRUE ~ NA_character_ 
                        ), levels = c("immediate", "root")),
         cd = factor(case_when(response=="RC_IC" ~  "continuous",
                        response=="RC" ~     "continuous",
                        response=="RD_ID" ~  "delayed",
                        response=="RD" ~     "delayed",
                        response=="IC" ~     "continuous",
                        response=="ID" ~     "delayed",
                        response=="otherC" ~     "continuous",
                        response=="otherD" ~     "delayed",
                        TRUE ~ NA_character_ 
                        ), levels = c("delayed", "continuous")),
         length = factor(length, levels = c("short", "medium", "long")),
         state_or_event= factor(state_or_event, levels = c("event", "state")),
        )

d1_stats1ri <- d1_stats1 %>% 
  drop_na(ri)

d1_stats1cd <- d1_stats1 %>% 
  drop_na(cd)

@@

fit1_ri <- brm( ri ~ state_or_event + length + (1|participantnum), family=bernoulli(), data = d1_stats1ri, chains = nchain, iter = niter )

>>>>

d1_stats1 <- d1 %>% 
  mutate(ri = factor(case_when(response=="RC_IC" ~  NA_character_,
                        response=="RC" ~     "root",
                        response=="RD_ID" ~  NA_character_,
                        response=="RD" ~     "root",
                        response=="IC" ~     "immediate",
                        response=="ID" ~     "immediate",
                        TRUE ~ NA_character_ 
                        ), levels = c("immediate", "root")),
         cd = factor(case_when(response=="RC_IC" ~  "continuous",
                        response=="RC" ~     "continuous",
                        response=="RD_ID" ~  "delayed",
                        response=="RD" ~     "delayed",
                        response=="IC" ~     "continuous",
                        response=="ID" ~     "delayed",
                        response=="otherC" ~     "continuous",
                        response=="otherD" ~     "delayed",
                        TRUE ~ NA_character_ 
                        ), levels = c("delayed", "continuous")),
         length = factor(length, levels = c("short", "medium", "long")),
         state_or_event = factor(state_or_event, levels = c("event", "state")),
         position = factor(case_when(is.na(position) ~  "middle",
                                     TRUE ~ position
                        ), levels = c("top", "bottom", "middle")),
         delay = factor(case_when(is.na(delay) ~  "none",
                                     TRUE ~ as.character(delay) 
                        ), levels = c("none", "2", "4", "6"))
        )

d1_stats1ri <- d1_stats1 %>% 
  drop_na(ri) %>% 
  drop_na(state_or_event) # drop chain

d1_stats1cd <- d1_stats1 %>% 
  drop_na(cd) %>% 
  drop_na(state_or_event)  # drop chain

@@

fit1_ri <- brm( ri ~ state_or_event + length + (1|participantnum), family=bernoulli(), data = d1_stats1ri, chains = nchain, iter = niter, control = list(adapt_delta = 0.99) )

### 2021-05-05 docs/resources/CogSci_paper/paper.tex
revised entire paper

Identifying the cause of an event (e.g.\ the stopping of an elevator) requires a judgment of actual causation, also known as singular or token causation~\cite{danks17}. Psychologists and philosophers have explored these judgments in detail~and have developed formal models of actual causation, including models based on Bayesian networks \cite{halpern16}, force dynamics \cite{wolfft17} and mental simulation~\cite{gerstenberg21}.  Here we present and evaluate an account of actual causation that highlights the role of temporal continuity.  On this account, the tripping of relay Q29BD is the main cause of the elevator stopping because the tripping initiated a continuous sequence of state changes that culminated in the stopping of the elevator.

Most accounts of actual causation are consistent with one of two broad views of causation: the counterfactual approach and the physical process approach~\cite{hall04}. The counterfactual approach suggests that the tripping of relay Q29BD is a cause of the elevator stopping because if the relay had not tripped then the elevator would not have stopped. The physical process view suggests that the relay tripping is a cause of the elevator stopping because the two are linked by a physical process involving the transmission of force or energy.  Both accounts need additional machinery in order to specify which among the many causes of an event is singled out as the main cause. Our continuity account fits most naturally with the physical process approach, and can be viewed as an attempt to bring out some implications of this approach for actual causation.

The continuity account relies on two core principles. First, the effect to be explained and its cause are both changes of state. The tripping of a relay qualifies as a candidate cause, but the steady-state setting of a relay does not. Our emphasis on state changes is consistent with the common view that causal relationships are relationships between events rather than facts or states of affairs, but  \citeA{glymourstone} point out that most Bayes net accounts of actual causation ignore changes of state. State changes are embedded in continuous time, and a change that begins at a specific moment is typically the direct result of an event that occurred an instant before. Focusing on state changes therefore motivates a principle of temporal continuity that allows continuous causal sequences to be traced back in time from an effect to its main cause. Some previous work on actual causation highlights the importance of time~\cite{stephanmw20e}, and our approach is highly compatible with work by Michotte and others suggesting that temporal information is often critical for identifying causes~\cite{youngs09,davisbr20}.  To our knowledge, however,  previous work has not explored the implications of temporal continuity in the way that we do here.

@@

\begin{figure*}
\captionsetup[subfigure]{labelformat=empty}
\begin{center}
\null\hfill
\subfloat[]{\includegraphics{figures/fig1a.eps}\label{subfig:1a}}
\hfill
\subfloat[]{\includegraphics{figures/fig1b.eps}\label{subfig:1b}}
\hfill\null
\end{center}
\caption{(a)(i) A network of particle detectors. All components are initially inactive. (ii) Detector \ev{R}{D} activates and activity propagates along the top branch. (iii) Detector \ev{R}{C} activates and ultimately triggers the activation of the GCM, which activates only when both  \ev{I}{D} and \ev{I}{C} are active. (b) An activation sequence in which (i) the top branch activates first, (ii) activation starts and finishes along the bottom branch while activation continues along the top branch, and (iii) the top branch completes its activation and ultimately triggers the GCM.}
%\caption{(a)(i) A smoke alarm connected to a chain of ionization detectors (top branch) and a chain of photoelectric detectors (bottom branch). All components are initially inactive (ii) Detector \ev{R}{D} activates and activity propagates along the top branch. (iii) Detector \ev{R}{C} activates and ultimately triggers the activation of the alarm, which activates only when both  \ev{I}{D} and \ev{I}{C} are active. (b) An activation sequence in which (i) the top branch activates first, (ii) activation starts and finishes along the bottom branch while activation continues along the top branch, and (iii) the top branch completes its activation and ultimately triggers the smoke alarm.}
\label{fig:intro}
\end{figure*}

@@

Although we focus on processes unfolding in continuous time, it will be convenient to divide up the temporal dimension into intervals brief enough that at most one event occurs per interval. The effect to be explained is an event that happens within one of these intervals: for example, the GCM's change of state from inactive to active. Given this setup, our theory can be formulated using a procedure that starts with the effect and steps backwards through time until the actual cause of the effect is identified.

The immediate cause of the GCM's activation must be an event that took place in the interval preceding the activation. In Figure~\ref{fig:intro}a, the event immediately preceding the GCM's activation was the activation of \ev{I}{C}. Having identified this immediate cause, we then step backwards and identify the immediate cause of this event, and so on.  The procedure terminates once we arrive at an event that has no immediate cause within the system of interest, 
and this cause is declared to be the main cause of the effect.  In Figure~\ref{fig:intro}a the main cause is the activation of \ev{R}{C}.

In most cases, the backward-tracing procedure just described will identify a single main cause of the effect. If the effect has no immediate cause within the system, then no main cause will be identified. Because there is at most one event per interval, the procedure can never identify more than one main cause.

@@

The boundaries of the causal system under consideration will depend on context. For simplicity, our discussion of Figure~\ref{fig:intro}a has focused only on events internal to the particle detection network, and has attemped to characterize which of these events is best viewed as the main cause of the GCM's activation. The particle detection network, however, could also be considered part of a broader causal system that includes both the network and the network's surroundings. For example, if the network is embedded in an apparatus for carbon dating, then the spontaneous decay of a carbon atom could be identified as the immediate cause of \ev{R}{C} and the main cause of the GCM's activation.

@@

@@

Although \ev{R}{D}, \ev{I}{D}, \ev{R}{C} and \ev{I}{C} all qualify as causes of the effect, Figure~\ref{fig:expt1} suggests that \ev{R}{C} tends to be singled out as the main cause. This result can be separated into two general conclusions.  First,  participants were more likely to choose root causes than immediate causes. Root and immediate causes were identical for the short sequences and therefore cannot be distinguished, but the results for medium and long sequences reveal a preference for root causes.  The second conclusion is that for dual branch sequences, the root cause on the continuous branch (\ev{R}{C}) is preferred to the root cause on the delayed branch (\ev{R}{D}). This result is consistent with the continuity account but also consistent with the probability raising account, because \ev{R}{C} guarantees the occurrence of the effect but \ev{R}{D} does not.

To support these two conclusions we ran two Bayesian mixed effects regression models using the \emph{brms} package in R~\cite{brms} with default priors\footnote{By default brms uses improper flat priors on the coefficients of fixed effects, and a half student-t prior with 3 degrees of freedom, a location of 0 and a scale of 2.5 on the intercept and on the standard deviations of the random effects.}. For the first model, each response was coded as root or immediate, and all responses that could not be classified in this way (including all responses for short sequences) were discarded. We then ran a logistic regression in which stativity (i.e.\ state vs event) and network length (long vs medium) were included as predictors of a binary dependent variable (root vs immediate). We included a random intercept for participant to allow for statistical dependencies between multiple responses from the same participant. The posterior mean of the intercept variable was 10.8 and the 95\% credible interval was $[2.4, 26.8]$, which supports the conclusion that root causes were chosen more often than immediate causes. The credible intervals for the other coefficients both included 0 (state: -2.7, $[-9.2, 1.2]$ ; long: 2.1, $[-1.7, 9.1]$), suggesting that the preference for root causes was not strongly affected by either stativity or network length.

The second analysis was very similar except that the binary dependent variable now indicated whether participants chose a cause on the continuous branch or the delayed branch. The posterior mean of the intercept variable was 1.7 but the 95\% credible interval included zero ($[-0.07, 3.8]$), suggesting only weak support for the conclusion that the continuous branch was chosen more often than the delayed branch. The credible intervals for the state variable excluded 0 (1.0, $[0.05, 1.9]$), suggesting that state sequences led to a stronger preference for the continuous branch than did event sequences.    Credible intervals for both length variables included 0 (medium: 0.4, $[-0.5,1.4]$ ; long: 0.5, $[-0.5,1.4]$, where the reference level is small), suggesting that the preference for the continuous branch was not strongly affected by network length.

@@

\begin{figure}
\begin{center}
%\includegraphics[width=8cm]{results_E1}
%\includegraphics{figures/expt1.eps}
\includegraphics{figures/expt1test.eps}
\end{center}
\caption{Experiment 1. (a) Order of the four key events for dual branch sequences. The probability raising and continuity accounts both single out \ev{R}{C} as the main cause. (b)  Responses for dual branch sequences. The delayed branch was either inactive (event sequences) or active (state sequences) at the start of the sequence. Error bars show the standard error of the mean estimated using 100 bootstrap samples.} 
\label{fig:expt1}
\end{figure}

\begin{figure}[t]
\begin{center}
%\includegraphics[width=8cm]{results_E2}
%\includegraphics{figures/expt2.eps}
\includegraphics{figures/expt2test.eps}
%\includegraphics{figures/results_E2.eps}
\end{center}
\caption{Experiment 2. (a) The probability raising and continuity accounts make different predictions for dual branch sequences in which \ev{R}{C} occurs before \ev{R}{D}. (b) The two accounts agree when \ev{R}{C} occurs after \ev{R}{D}.
(c) Responses for the three possible orders of the key events.} 
\label{fig:expt2}
\end{figure}

@@

The results of Experiment 1 are broadly consistent with both the continuity and probability raising accounts, and the goal of Experiment 2 was to distinguish between these accounts. The two make different predictions for dual branch sequences in which the delayed branch activates after activation has already begun along the continuous branch. One such sequence is shown in Figure~\ref{fig:intro}b. 
In cases like this, the continuity account still treats \ev{R}{C} as the primary cause, but the probability raising account now treats \ev{R}{D} as the primary cause. To see why, let $b$ (for baserate) be the probability of activation along along a given branch during a given trial. At the start of a dual branch trial, the probability that the GCM will activate is $b^2$, because independent activations are required along both branches. After \ev{R}{C} the probability increases to $b$, and after \ev{R}{D} the probability increases to 1. The two probability increments are therefore $b - b^2$ and $1 - b$, and it is straightforward to show that $b - b^2 \le 1 - b$ for all values of $b$, with equality attained only when $b = 1$.  As a result, the probability raising account chooses \ev{R}{D} for virtually all values of the baserate, and if $b=1$ the account is indifferent between \ev{R}{C} and \ev{R}{D}.

The materials and procedure for Experiment 2 are similar to those for Experiment 1, and we will highlight only the few points of difference.  As for Experiment 1, no explicit information about the baserate $b$ was provided. For each network shown, activation eventually occurred along each branch, and some participants may have therefore assumed that $b=1$. Most participants, however, probably implicitly assumed that $b < 1$, which allows for the possibility of observation periods during which no activations occurred. We therefore assume that $b < 1$ when discussing the predictions of the probability raising account.

@@

To further analyze the data we used mixed effects models analogous to the two described for Experiment 1. The first analysis used a logistic regression in which order (i.e.\ the temporal order of the key events) was included as a predictor of a binary dependent variable (root or immediate).  The posterior mean of the intercept variable was 5.2 and the 95\% credible interval $[2.0, 9.4]$, which supports the conclusion that root causes were chosen more often than immediate causes. The credible intervals for both order variables included 0 (\{\ev{R}{D},\ev{R}{C},\ev{I}{D}\}: -0.2, $[-1.4,1.2]$; \{\ev{R}{D},\ev{I}{D},\ev{R}{C}\}: -0.4 $[-1.7,0.9]$, where the reference level was \{\ev{R}{C},\ev{R}{D},\ev{I}{D}\}), suggesting that the preference for root causes was fairly consistent across the three orders.

The second and more critical analysis used a binary dependent variable that indicated whether participants chose a cause on the continuous branch or the delayed branch.  The posterior mean of the intercept variable was 1.9 (95\% credible interval $[1.3, 2.5]$), indicating that the continuous branch was chosen more often than the delayed branch.  The credible intervals for both order variables included 0 ( \{\ev{R}{D},\ev{R}{C},\ev{I}{D}\}: -0.4, $[-1,0.2]$ ; \{\ev{R}{D},\ev{I}{D},\ev{R}{C}\}: -0.5, $[-1,0.1]$) suggesting that the preference for the continuous branch was of similar strength across all three orders.

@@

We presented an account of actual causation that highlights the role of temporal continuity and described experiments that support two of its predictions. First, people tend to identify root causes rather than immediate causes as the main cause of an effect. Second, when an effect is produced by the convergence of multiple causal pathways, participants tend to identify the pathway that has temporal continuity with the effect as the primary cause.

Because the continuity account considers changes of state in continuous time, it can exploit a ``no coincidences'' principle to identify a single main cause of an effect. This principle applies broadly to causation in physical systems, but is less applicable to social scenarios such as voting scenarios~\cite{livengood13}. Exceptions to the ``no coincidences'' principle are possible even for physical systems, and the continuity account can capture the illusory perceptions of actual causation that arise in some such cases~\cite{thorstadw16}. For example, an 11 year old boy whacked a stick against a telephone pole at the instant at which the Great Northeastern Blackout hit his town, and ran home terrified that he had caused the power outage~\cite{gelbr1965night}.

@@

As suggested earlier, standard formulations of the physical process view treat both \ev{R}{C} and \ev{R}{D} as causes of the effect in Figure~\ref{fig:intro}. Our data are consistent with the possibility that people view \ev{R}{C} and \ev{R}{D} as equally valid causes, but pick \ev{R}{C} when forced to break the tie for reasons that may be relatively superficial. For example, perhaps \ev{R}{C} is preferred because the activation on the continuous branch is continuous both in space and time, and therefore more prototypical of a causal process than the activation along the delay branch. We have planned a future study that asks whether the \ev{R}{C} over \ev{R}{D} is fundamental or superficial using a task in which participants rank multiple candidate causes in order of importance, and in which ties are permitted.

@@

AF carried out this work at the University of Melbourne while enrolled in a Master's degree at the \'{E}cole Normale Sup\'{e}rieure. 
% add financial support?

>>>>

Identifying the primary cause of an event (e.g.\ the stopping of an elevator) requires a judgment of actual causation, also known as singular or token causation~\cite{danks17}. Psychologists and philosophers have explored actual causation judgments in detail~and have developed formal models of actual causation, including models based on Bayesian networks \cite{halpern16}, force dynamics \cite{wolfft17} and mental simulation~\cite{gerstenberg21}.  Here we present and evaluate an account of actual causation that highlights the role of temporal continuity.  On this account, the tripping of relay Q29BD is the primary cause of the elevator stopping because the tripping initiated a continuous sequence of state changes that culminated in the stopping of the elevator. Identifying this primary cause may involve two steps: the first is to assemble a set of causes (e.g.\ a set that includes all of the state changes just mentioned), and the second is to select the primary cause from among this set~\cite{stephanmw20e}. Our general approach has implications for both steps, but in our setting the second step (often called causal selection) is the more challenging of the two, and we focus on it throughout.

Most accounts of actual causation are consistent with one of two broad views of causation: the counterfactual approach and the physical process approach~\cite{hall04}. The counterfactual approach suggests that the tripping of relay Q29BD is a cause of the elevator stopping because if the relay had not tripped then the elevator would not have stopped. The physical process view suggests that the relay tripping is a cause of the elevator stopping because the two are linked by a physical process involving the transmission of force or energy.  Both accounts need additional machinery in order to specify which among the many causes of an event is singled out as the primary cause. Our continuity account fits most naturally with the physical process approach, and can be viewed as an attempt to bring out some implications of this approach for actual causation.

The continuity account relies on two core principles. First, the effect to be explained and its cause are both changes of state. The tripping of a relay qualifies as a candidate cause, but the steady-state setting of a relay does not. Our emphasis on state changes is consistent with the common view that causal relationships are relationships between events rather than facts or states of affairs, but  \citeA{glymourstone} point out that most Bayes net accounts of actual causation ignore changes of state. State changes are embedded in continuous time, and a change that begins at a specific moment is typically the direct result of an event that occurred an instant before. Focusing on state changes therefore motivates a principle of temporal continuity that allows continuous causal sequences to be traced back in time from an effect to its primary cause. Some previous work on actual causation highlights the importance of time~\cite{stephanmw20e}, and our approach is highly compatible with work by Michotte and others suggesting that temporal information is often critical for identifying causes~\cite{youngs09,davisbr20}.  To our knowledge, however,  previous work has not explored the implications of temporal continuity in the way that we do here.

@@

\begin{figure*}
\captionsetup[subfigure]{labelformat=empty}
\begin{center}
\null\hfill
\subfloat[]{\includegraphics{figures/fig1a.eps}\label{subfig:1a}}
\hfill
\subfloat[]{\includegraphics{figures/fig1b.eps}\label{subfig:1b}}
\hfill\null
\end{center}
\vspace{-0.3in}
\caption{(a)(i) A network of particle detectors. All components are initially inactive. (ii) Detector \ev{R}{D} activates and activity propagates along the top branch. (iii) Detector \ev{R}{C} activates and ultimately triggers the activation of the GCM, which activates only when both  \ev{I}{D} and \ev{I}{C} are active. (b) An activation sequence in which (i) the top branch activates first, (ii) activation starts and finishes along the bottom branch while activation continues along the top branch, and (iii) the top branch completes its activation and ultimately triggers the GCM.}
%\caption{(a)(i) A smoke alarm connected to a chain of ionization detectors (top branch) and a chain of photoelectric detectors (bottom branch). All components are initially inactive (ii) Detector \ev{R}{D} activates and activity propagates along the top branch. (iii) Detector \ev{R}{C} activates and ultimately triggers the activation of the alarm, which activates only when both  \ev{I}{D} and \ev{I}{C} are active. (b) An activation sequence in which (i) the top branch activates first, (ii) activation starts and finishes along the bottom branch while activation continues along the top branch, and (iii) the top branch completes its activation and ultimately triggers the smoke alarm.}
\label{fig:intro}
\end{figure*}

@@

Although we focus on processes unfolding in continuous time, it will be convenient to divide up the temporal dimension into intervals brief enough that at most one event occurs per interval. The effect to be explained is an event that happens within one of these intervals: for example, the GCM's change of state from inactive to active. The continuity account suggests that there can be at most one event that triggered a continuous sequence of state changes that culminated in the effect, and proposes that this triggering event (if it exists) is singled out as the primary cause. The triggering event can be identified using a procedure that starts with the effect and steps backwards through time until the primary cause is identified.

The immediate cause of the GCM's activation must be an event that took place in the interval preceding the activation. In Figure~\ref{fig:intro}a, the event immediately preceding the GCM's activation was the activation of \ev{I}{C}. Having identified this immediate cause, we then step backwards and identify the immediate cause of this event, and so on.  The procedure terminates once we arrive at an event that has no immediate cause within the system of interest, 
and this cause is selected as the primary cause of the effect.  In Figure~\ref{fig:intro}a the primary cause is the activation of \ev{R}{C}.

In most cases, the backward-tracing procedure just described will identify a single primary cause of the effect. If the effect has no immediate cause within the system, then no primary cause will be identified. Because there is at most one event per interval, the procedure can never identify more than one primary cause.

@@

The boundaries of the causal system under consideration will depend on context. For simplicity, our discussion of Figure~\ref{fig:intro}a has focused only on events internal to the particle detection network, and has attemped to characterize which of these events is best viewed as the primary cause of the GCM's activation. The particle detection network, however, could also be considered part of a broader causal system that includes both the network and the network's surroundings. For example, if the network is embedded in an apparatus for carbon dating, then the spontaneous decay of a carbon atom could be identified as the immediate cause of \ev{R}{C} and the primary cause of the GCM's activation.

@@

To apply the probability raising account to the dual branch stimuli, let $b$ (for base rate) be the probability of activation along along a given branch during a given trial. At the start of a dual branch trial, the probability that the GCM will activate is $b^2$, because independent activations are required along both branches. After \ev{R}{D} the probability increases to $b$, and after \ev{R}{C} the probability increases to 1. The two probability increments are therefore $b - b^2$ and $1 - b$, and it is straightforward to show that $b - b^2 \le 1 - b$ for all $b$, with equality attained only when $b = 1$.  As a result, the probability raising account selects \ev{R}{C} if $b<1$, and is indifferent between \ev{R}{C} and \ev{R}{D} if $b = 1$.

For each dual branch network shown, activation eventually occurred along each branch, and some participants may have therefore assumed that $b=1$. Most participants, however, probably implicitly assumed that $b < 1$, which allows for the possibility of trials during which a branch never activated. We will therefore assume that $b < 1$ when considering the predictions of the probability raising account.

@@

Although \ev{R}{D}, \ev{I}{D}, \ev{R}{C} and \ev{I}{C} all qualify as causes of the effect, Figure~\ref{fig:expt1} suggests that \ev{R}{C} tends to be singled out as the primary cause. This result can be separated into two general conclusions.  First,  participants were more likely to choose root causes than immediate causes. Root and immediate causes were identical for the short sequences and therefore cannot be distinguished, but the results for medium and long sequences reveal a preference for root causes.  The second conclusion is that for dual branch sequences, the root cause on the continuous branch (\ev{R}{C}) is preferred to the root cause on the delayed branch (\ev{R}{D}). As explained earlier, this result is consistent with both the continuity and probability raising accounts.

To support these two conclusions we ran two Bayesian mixed effects regression models using the \emph{brms} package in R~\cite{brms} with default priors.\footnote{By default brms uses improper flat priors on the coefficients of fixed effects, and a half student-t prior with 3 degrees of freedom, a location of 0 and a scale of 2.5 on the intercept and on the standard deviations of the random effects.} For the first model, each response was coded as root or immediate, and all responses that could not be classified in this way (including all responses for short sequences) were discarded. We then ran a logistic regression in which stativity (i.e.\ state vs event) and network length (long vs medium) were included as predictors of a binary dependent variable (root vs immediate). We included a random intercept for participant to allow for statistical dependencies between multiple responses from the same participant. The posterior mean of the intercept variable was 11.0 and the 95\% credible interval was $[2.6, 28.7]$, which supports the conclusion that root causes were chosen more often than immediate causes. The credible intervals for the other coefficients both included 0 (state: -2.7, $[-9.3, 1.1]$ ; long: 2.1, $[-1.7, 9.0]$), suggesting that the preference for root causes was not strongly affected by either stativity or network length.

The second analysis was very similar except that the binary dependent variable now indicated whether participants chose a cause on the continuous branch or the delayed branch. The posterior mean of the intercept variable was 1.7 but the 95\% credible interval included zero ($[-0.06, 3.7]$), suggesting only weak support for the conclusion that the continuous branch was chosen more often than the delayed branch. The credible intervals for the state variable excluded 0 (1.0, $[0.05, 1.9]$), suggesting that state sequences led to a stronger preference for the continuous branch than did event sequences.    Credible intervals for both length variables included 0 (medium: 0.4, $[-0.5,1.4]$ ; long: 0.5, $[-0.5,1.4]$, where the reference level is small), suggesting that the preference for the continuous branch was not strongly affected by network length.

@@

\begin{figure}
\begin{center}
%\includegraphics[width=8cm]{results_E1}
%\includegraphics{figures/expt1.eps}
\includegraphics{figures/expt1test.eps}
\end{center}
\vspace{-0.1in}
\caption{Experiment 1. (a) Order of the four key events for dual branch sequences. The probability raising and continuity accounts both single out \ev{R}{C} as the primary cause. (b)  Responses for dual branch sequences. The delayed branch was either inactive (event sequences) or active (state sequences) at the start of the sequence. Error bars show the standard error of the mean estimated using 100 bootstrap samples.} 
\label{fig:expt1}
\end{figure}

\begin{figure}[t]
\begin{center}
%\includegraphics[width=8cm]{results_E2}
%\includegraphics{figures/expt2.eps}
\includegraphics{figures/expt2test.eps}
%\includegraphics{figures/results_E2.eps}
\end{center}
\vspace{-0.1in}
\caption{Experiment 2. (a) The probability raising and continuity accounts make different predictions for dual branch sequences in which \ev{R}{C} occurs before \ev{R}{D}. (b) The two accounts agree when \ev{R}{C} occurs after \ev{R}{D}.
(c) Responses for the three possible orders of the key events.} 
\label{fig:expt2}
\end{figure}

@@

The results of Experiment 1 are broadly consistent with both the continuity and probability raising accounts, and the goal of Experiment 2 was to distinguish between these accounts. The two make different predictions for dual branch sequences in which the delayed branch activates after activation has already begun along the continuous branch. One such sequence is shown in Figure~\ref{fig:intro}b. 
In cases like this, the continuity account still treats \ev{R}{C} as the primary cause, but the probability raising account now treats \ev{R}{D} as the primary cause. For dual branch stimuli, the argument presented earlier shows that the probability raising account identifies the second of the two root causes (i.e.\ \ev{R}{C} in Experiment 1 and \ev{R}{D} for the sequence in Figure~\ref{fig:intro}b) as the primary cause if the base rate $b < 1$.  If $b=1$ the account is unable to select any event as the primary cause.

The materials and procedure for Experiment 2 are similar to those for Experiment 1, and we will highlight only the few points of difference.

@@

To further analyze the data we used mixed effects models analogous to the two described for Experiment 1. The first analysis used a logistic regression in which order (i.e.\ the temporal order of the key events) was included as a predictor of a binary dependent variable (root or immediate).  The posterior mean of the intercept variable was 5.0 and the 95\% credible interval $[2.0, 8.9]$, which supports the conclusion that root causes were chosen more often than immediate causes. The credible intervals for both order variables included 0 (\{\ev{R}{D},\ev{R}{C},\ev{I}{D}\}: -0.2, $[-1.5,1.2]$; \{\ev{R}{D},\ev{I}{D},\ev{R}{C}\}: -0.4 $[-1.7,0.9]$, where the reference level was \{\ev{R}{C},\ev{R}{D},\ev{I}{D}\}), suggesting that the preference for root causes was fairly consistent across the three orders.

The second and more critical analysis used a binary dependent variable that indicated whether participants chose a cause on the continuous branch or the delayed branch.  The posterior mean of the intercept variable was 1.9 (95\% credible interval $[1.3, 2.5]$), indicating that the continuous branch was chosen more often than the delayed branch.  The credible intervals for both order variables included 0 ( \{\ev{R}{D},\ev{R}{C},\ev{I}{D}\}: -0.4, $[-0.9,0.1]$ ; \{\ev{R}{D},\ev{I}{D},\ev{R}{C}\}: -0.5, $[-1,0.1]$) suggesting that the preference for the continuous branch was of similar strength across all three orders.

@@

We presented an account of actual causation that highlights the role of temporal continuity and described experiments that support two of its predictions. First, people tend to identify root causes rather than immediate causes as the primary cause of an effect. Second, when an effect is produced by the convergence of multiple causal pathways, participants tend to identify the pathway that has temporal continuity with the effect as the primary cause.

Because the continuity account considers changes of state in continuous time, it can exploit a ``no coincidences'' principle to identify a single primary cause of an effect. This principle applies broadly to causation in physical systems, but is less applicable to social scenarios such as voting scenarios~\cite{livengood13}. Exceptions to the ``no coincidences'' principle are possible even for physical systems, and the continuity account can capture the illusory perceptions of actual causation that arise in some such cases~\cite{thorstadw16}. For example, an 11 year old boy whacked a stick against a telephone pole at the instant at which the Great Northeastern Blackout hit his town, and ran home terrified that he had caused the power outage~\cite{gelbr1965night}.

@@

As suggested earlier, standard formulations of the physical process view treat both \ev{R}{C} and \ev{R}{D} as causes of the effect in Figure~\ref{fig:intro}. Our data are consistent with the possibility that people view \ev{R}{C} and \ev{R}{D} as equally valid causes, but pick \ev{R}{C} when forced to break the tie for reasons that may be relatively superficial. For example, perhaps \ev{R}{C} is preferred because the activation on the continuous branch is continuous both in space and time, and therefore more prototypical of a causal process than the activation along the delay branch. We have planned a future study that asks whether the preference for \ev{R}{C} over \ev{R}{D} is fundamental or superficial using a task in which participants rank multiple candidate causes in order of importance, and in which ties are permitted.

@@

AF carried out this work at the University of Melbourne as part of a Master's degree at the \'{E}cole Normale Sup\'{e}rieure. 
% add financial support?
